---
title: "Cluster Analysis of Milwaukee Schools"
logo: logo.png
toc: true
output: 
  pdf_document:
    includes:
      in_header: header.tex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(scales)
library(wisconsink12)
library(cityforwardcollective)
library(sf)
library(glue)
library(ggtext)
library(cluster)
library(purrr)

set.seed(1234)

geo_schools_df <- read_csv("../../000_data_temp/geocoded_mke_schools.csv") %>%
  select(school_year, dpi_true_id, lat, long)

geo_schools <- geo_schools_df %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326)

sbds <- st_read("../../Shapefiles/Milwaukee/MPS_SBD_2022/MPS_B_Boundaries.shp") %>%
  st_transform(., crs = st_crs(4326)) %>%
  group_by(DISTRICT) %>%
  summarise()

city_limits <- st_read("../../Shapefiles/Milwaukee/City Limits/citylimit.shp") %>%
  st_transform(., crs = st_crs(4326))


sbd_schools <- st_intersection(geo_schools, sbds)

no_geo <- sbd_schools %>%
  as_tibble() %>%
  select(school_year,
         dpi_true_id,
         sbd = DISTRICT)

mke_rc <- make_mke_rc() %>%
  left_join(., geo_schools_df)

demo <- mke_rc %>%
  filter(school_year == max(report_cards$school_year) & !is.na(per_am_in)) %>% 
  select(per_am_in:per_lep) %>%
  select(-c(per_am_in, per_nh_opi)) %>%
  mutate_all(replace_na, replace = 0)

demo_geo <- mke_rc %>%
  filter(school_year == max(report_cards$school_year) & !is.na(per_am_in)) %>% 
  select(per_am_in:per_lep,
         lat, long) %>%
  select(-c(per_am_in, per_nh_opi)) %>%
  mutate_all(replace_na, replace = 0)

scaled_demo <- scale(demo)

dist_demo <- dist(scaled_demo)

scaled_demo_geo <- scale(demo_geo)

dist_demo_geo <- dist(scaled_demo_geo)

k_clust <- kmeans(dist_demo, centers = 3, iter.max = 1000, nstart = 100)
k_clust_geo <- kmeans(dist_demo_geo, centers = 3, iter.max = 1000, nstart = 100)


ach_clustered <- mke_rc %>%
  filter(school_year == max(report_cards$school_year) & !is.na(per_am_in)) %>%
  mutate(cluster = k_clust$cluster) %>%
  left_join(., no_geo) 

ach_clustered_geo <- mke_rc %>%
  filter(school_year == max(report_cards$school_year) & !is.na(per_am_in)) %>%
  mutate(cluster = k_clust_geo$cluster) %>%
  left_join(., no_geo) %>%
  mutate(cluster = case_when(cluster == 1 ~ 3,
                             cluster == 2 ~ 1,
                             cluster == 3 ~ 2))


c_demo <- ach_clustered %>%
  group_by(cluster) %>%
  summarise_at(vars(per_am_in:per_lep, lat, long, overall_score, sch_ach, sch_growth), .funs = mean, na.rm = TRUE) %>%
  modify_at(vars(per_am_in:per_lep), scales::percent) %>%
  ungroup()


c_demo_geo <- ach_clustered_geo %>%
  group_by(cluster) %>%
  summarise_at(vars(per_am_in:per_lep, lat, long, overall_score, sch_ach, sch_growth), .funs = mean, na.rm = TRUE) %>%
  modify_at(vars(per_am_in:per_lep), scales::percent)
```

\newpage
# Introduction

## Purpose

The purpose of this analysis is to cluster Milwaukee K12 schools using a k-means clustering algorithm. We will evaluate two clustering methods:

* Method A will not account for the physical location of the school 
* Method B will account for the physical location of the school

Method A was selected because it was simpler and because Method B only served to nudge edge-case schools into or away from physically nearby clusters depending on how similar they were.

## Key Findings

The results are three clusters, here called Cluster 1, Cluster 2, and Cluster 3. 

* **Cluster 1** is the second largest cluster, and it serves second most disadvantaged student body, located primarily on the south side of the city.
* **Cluster 2** is the smallest cluster, and it is by far the least disadvantaged student body.
* **Cluster 3** is the largest and most disadvantaged cluster; located primarily on the north side and predominantly serving Black students, it also has highest proportion of students with disabilities and economically disadvantaged students.

Below are summary tables that provide student body demographics, School Report Card outcomes for the 2020-21 school year, and the count of schools in each School Board District (SBD).^[This report uses School Board District boundaries adopted by MPS in 2022.]

```{r}
c_demo %>%
  select(cluster, per_b_aa, per_hisp_lat, per_white, per_asian, per_swd, per_lep, per_ed, lat, long) %>%
  left_join(., ach_clustered %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Black" = per_b_aa,
         "Hisp/Lat" = per_hisp_lat,
         "White" = per_white,
         "Asian" = per_asian,
         "SwD" = per_swd,
         "LEP" = per_lep,
         "ECD" = per_ed) %>%
  kable(booktabs = T, align = "c") %>%
  kable_styling(latex_options = "hold_position")
```

```{r}
c_demo %>%
  select(cluster, overall_score, sch_ach, sch_growth) %>%
  left_join(., ach_clustered %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Overall Score" = overall_score,
         "Achievement" = sch_ach,
         "Growth" = sch_growth) %>%
  modify_at(3:5, round, 1) %>%
  kable(booktabs = T, align = "c") %>%
  kable_styling(latex_options = "hold_position")
```

```{r}
sbd_clust <- ach_clustered %>%
  mutate(sbd = replace_na(sbd, "ZNone"),
         sbd = ifelse(sbd == "ZNone", sbd, paste0("SBD", sbd))) %>%
  group_by(cluster, sbd) %>%
  tally() %>%
  arrange(sbd) %>%
  pivot_wider(names_from = sbd, values_from = n, values_fill = 0)

sbd_clust %>%
  kbl(booktabs = T, 
      align = "c", col.names = c("Cluster", glue("SBD{1:8}"), "None")) %>%
  kable_styling(latex_options = "HOLD_position")

```


```{r fig.height = 7}
ach_clustered %>%
  mutate(cluster = as.factor(cluster)) %>%
  select(dpi_true_id,
         cluster) %>%
  left_join(., geo_schools) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(data = sbds, size = .1) +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT != 7), inherit.aes = FALSE,
               size = 5, color = "red") +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT == 7), inherit.aes = FALSE,
               size = 5, nudge_y = -.01,  color = "red") +
  geom_sf(aes(fill = cluster), alpha = .9, shape = 21) +
  theme_void() +
  theme(plot.caption = element_textbox(width = unit(5, "in"),
                                       hjust = 0)) +
  labs(fill = "Cluster",
       title = "Clustered School Locations",
       caption = glue("Boundaries and <span style='color:{'red'}'>**red labels**</span> ",
                      "represent School Board Districts adopted in 2022."))

```

```{r fig.height = 7}
l_clust <- ach_clustered %>%
  select(14:23, cluster, dpi_true_id) %>%
  select(-c(per_am_in, per_nh_opi)) %>%
  pivot_longer(cols = 1:8, names_to = "group", values_to = "value") %>%
  mutate(group = case_when(group == "per_white" ~ "White",
                           group == "per_tom" ~ "Two or more",
                           group == "per_swd" ~ "Students with Disabilities",
                           group == "per_lep" ~ "Limited English Proficiency",
                           group == "per_hisp_lat" ~ "Latino",
                           group == "per_ed" ~ "Economically Disadvantaged",
                           group == "per_b_aa" ~ "Black",
                           group == "per_asian" ~ "Asian"),
         group = factor(group, levels = rev(c("Asian",
                                          "Black",
                                          "Latino",
                                          "Two or more",
                                          "White",
                                          "Students with Disabilities",
                                          "Limited English Proficiency",
                                          "Economically Disadvantaged"))))

l_clust %>%
  ggplot(aes(group, value, color = as.character(cluster))) +
  geom_jitter(width = .2, alpha = .5, height = 0, size = 1) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.text.y = element_text(size = 10)) +
  scale_y_continuous(labels = percent) +
  scale_x_discrete(labels = function(x) str_wrap(x, 15)) +
  coord_flip() +
  theme(legend.position = "bottom",
        plot.title.position = "plot") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  labs(color = "Cluster",
       y = "Percent of School Enrollment",
       x = "Student Subgroup",
       title = "Distribution of School Demographics by Cluster")
```



\newpage
# Analysis and Methodology

## Data

The data used in this analysis is sourced primarily from the Wisconsin School Report Cards, though the physical school location is sourced from DPI School Directories.

## Clustering Variables

The variables used as inputs are those regarding the student body composition, i.e. racial subgroups, students with disabilities, English learner status, and economic status, all represented as a percentage of total school enrollment. Two racial subgroups were omitted from this analysis: 1) American Indian or Alaskan Native and 2) Native Hawaiian/Other Pacific Islander. These groups were omitted because they make up very small proportions of school enrollment citywide, so small differences had an outsized impact on clustering.

The following variables were included:

* Percent Asian 
* Percent Black 
* Percent Latino 
* Percent two or more races
* Percent white
* Percent students with disabilities
* Percent limited English proficiency
* Percent economically disadvantaged
* Latitude of school (Method B only)
* Longitude of school (Method B only)

## Number of Clusters

To determine the number of clusters, we will use the elbow and the silhouette methods. Both of these methods provide a way to evaluate how well clusters are made so we can compare the outcomes of different numbers of clusters.

The plot below shows the total within-cluster sum of squares (i.e. the intra-cluster variation) on the y axis, and the number of clusters on the x axis. We are looking to minimize the variation while keeping the number of clusters as small as we can. In other words, we want to find the point at which adding more clusters doesn't decrease variation enough to warrent adding that additional cluster. 

```{r out.width="75%", fig.align='center'}

tot_withinss <- map_dbl(1:10, function(k) {
  model = kmeans(x = dist_demo, centers = k)
  model$tot.withinss
})

elbow_df <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss
)

elbow_df %>%
  ggplot(aes(k, tot_withinss)) +
  geom_line() +
  geom_vline(xintercept = 3, linetype = 2, color = "red") +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Elbow method",
       x = "Number of Clusters",
       y = "Intra-Cluster Variation")
```

When examining the plot, the optimum number of clusters is observed as an  "elbow" in the line. For our case here, three clusters seems to be that point. Now let's see what the silhouette shows us.

The silhouette method will evaluate how well points fall within their clusters, and we are looking to maximize this measure, shown as the y axis on the plot below.

```{r out.width="75%", fig.align='center'}
sil_width <- map_dbl(2:10, function(k){
  model <- pam(x = dist_demo, k = k)
  model$silinfo$avg.width
})
sil_df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

sil_df %>%
  ggplot(aes(k, sil_width)) +
  geom_line() +
  geom_vline(xintercept = 3, linetype = 2, color = "red") +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Silhouette Method",
       x = "Number of Clusters",
       y = "Average Silhouette Width")
```

Based on this method, the optimal number of clusters is also three. **Therefore, we will settle on three clusters.**

\newpage
## Method A: Clustering without Geography

Method A clusters schools based on characteristics of the student body, i.e. race/ethnicity, disability status, English learner status, and economic status. This method does not directly account for the location of the school within the city, though because Milwaukee is so segregated, we will undoubtedly see that our clusters are grouped geographically.


```{r}
c_demo %>%
  select(cluster, per_b_aa, per_hisp_lat, per_white, per_asian, per_swd, per_lep, per_ed, lat, long) %>%
  left_join(., ach_clustered %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Black" = per_b_aa,
         "Hisp/Lat" = per_hisp_lat,
         "White" = per_white,
         "Asian" = per_asian,
         "SwD" = per_swd,
         "LEP" = per_lep,
         "ECD" = per_ed) %>%
  kable(booktabs = T, caption = "Method A Cluster Descriptions",
        align = "c") %>%
  kable_styling(latex_options = "hold_position")
```

Table 1 shows the average demographics of the schools within each cluster, along with the count of schools. 

* **Cluster 1** is comprised of schools where Hispanic/Latino students make up the largest share of the student body. The percent of Students with Disabilities and Economically Disadvantaged students is much higher than Cluster 2, though slightly lower than Cluster 3. The percent of students with Limited English Proficiency is highest among the three clusters.
* **Cluster 2** is comprised of more diverse schools, and the percent of Students with Disabilities and Economically Disadvantaged students is lowest among the three clusters.
* **Cluster 3** is comprised of schools where Black students make up the largest share of the student body, and the percent of Students with Disabilities and Economically Disadvantaged students is highest among the three clusters.

A more simplified way to interpret the clusters is that Cluster 3 is the most disadvantaged because we have observed over the years that Black students, Students with Disabilities, and Economically Disadvantaged students have the lowest access to high quality schools. Cluster 2 is slightly less disadvantaged, and Cluster 1 is by far the least disadvantaged.

```{r}
c_demo %>%
  select(cluster, overall_score, sch_ach, sch_growth) %>%
  left_join(., ach_clustered %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Overall Score" = overall_score,
         "Achievement" = sch_ach,
         "Growth" = sch_growth) %>%
  modify_at(3:5, round, 1) %>%
  kable(booktabs = T, caption = "Method A Cluster Report Card Scores (2020-21)",
        align = "c") %>%
  kable_styling(latex_options = "hold_position")
```

Table 2, showing average School Report Card scores for each cluster, supports these descriptions. Cluster 3 schools show the lowest scores, Cluster 1 scores are in the middle, and Cluster 2 shows the highest scores.

Finally, let's take a look at where the schools fall within the city, including their School Board District. Table 3 provides the count of schools from each cluster that fall within each School Board District (SBD).^[School Board Districts are those adopted by the MPS Board of Directors in 2022.] The following map plots the location of the school within the city, with SBD boundaries and text labels provided for reference. 

Taken together, we can see that Cluster 1 is concentrated on the south side of the city, Cluster 3 is concentrated on the north side, and Cluster 2 is fairly spread out across the city, though with an area of concentration in SBD8.

```{r}
sbd_clust <- ach_clustered %>%
  mutate(sbd = replace_na(sbd, "ZNone"),
         sbd = ifelse(sbd == "ZNone", sbd, paste0("SBD", sbd))) %>%
  group_by(cluster, sbd) %>%
  tally() %>%
  arrange(sbd) %>%
  pivot_wider(names_from = sbd, values_from = n, values_fill = 0)

sbd_clust %>%
  kbl(booktabs = T, caption = "Method A Cluster Location in School Board Districts",
      align = "c", col.names = c("Cluster", glue("SBD{1:8}"), "None")) %>%
  kable_styling(latex_options = "HOLD_position")

```

```{r fig.height = 7}
ach_clustered %>%
  mutate(cluster = as.factor(cluster)) %>%
  select(dpi_true_id,
         cluster) %>%
  left_join(., geo_schools) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(data = sbds, size = .1) +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT != 7), inherit.aes = FALSE,
               size = 5, color = "red") +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT == 7), inherit.aes = FALSE,
               size = 5, nudge_y = -.01,  color = "red") +
  geom_sf(aes(fill = cluster), alpha = .9, shape = 21) +
  theme_void() +
  theme(plot.caption = element_textbox(width = unit(5, "in"),
                                       hjust = 0)) +
  labs(fill = "Cluster",
       title = "Method A School Locations",
       caption = glue("Boundaries and <span style='color:{'red'}'>**red labels**</span> ",
                      "represent School Board Districts adopted in 2022."))

```

\newpage
## Method B: Clustering with Geography

Method B is essentially the same as Method A, except it adds the school's latitude and longitude as variables to the clustering algorithm, thereby accounting for the school's physical location. Reviewing the cluster descriptions in Table 4 below, we see that the clusters are quite similar to those of Method A. The broad interpretations we made above still hold -- Cluster 3 remains the most disadvantaged, Cluster 2 the least, and Cluster 1 in the middle.

```{r}
c_demo_geo %>%
  select(cluster, per_b_aa, per_hisp_lat, per_white, per_asian, per_swd, per_lep, per_ed, lat, long) %>%
  left_join(., ach_clustered_geo %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Black" = per_b_aa,
         "Hisp/Lat" = per_hisp_lat,
         "White" = per_white,
         "Asian" = per_asian,
         "SwD" = per_swd,
         "LEP" = per_lep,
         "ECD" = per_ed)  %>%
  kable(booktabs = T, caption = "Method B Cluster Descriptions") %>%
  kable_styling(latex_options = "hold_position")
```

Further, we observe the same trend in Report Card Scores, as shown in Table 5 below.

```{r}
c_demo_geo %>%
  select(cluster, overall_score, sch_ach, sch_growth) %>%
  left_join(., ach_clustered_geo %>%
                group_by(cluster) %>%
                summarise(N = n()), by = "cluster") %>%
  select("Cluster" = cluster,
         N,
         "Overall Score" = overall_score,
         "Achievement" = sch_ach,
         "Growth" = sch_growth) %>%
  modify_at(3:5, round, 1) %>%
  kbl(booktabs = T, caption = "Method B Cluster Report Card Scores (2020-21)",
        align = "c") %>%
  kable_styling(latex_options = "hold_position")
```

This is not surprising, since most schools were grouped in the same cluster as Method A. In fact, there were only six schools that had a different cluster in Method B than in Method A.


Reviewing the characteristics of these schools shown in Table 6, we can see that they are racially diverse, making them edge cases for any cluster. In fact, what seems to have happened is that the addition of geography to the algorithm has nudged the clustering in such a way that these edge cases were grouped more in line with schools in geographic proximity.

For instance, Divine Savior Holy Angles (DSHA) was placed in Cluster 1 with Method A, which is the predominantly south-side cluster. When geography is added with Method B, DSHA is clustered in with Cluster 2, the least disadvantaged/most geographically distributed cluster. In other words, When location was added to the mix, DSHA was deemed to be too far from Cluster 1, even though the student body composition would place it there.

Similarly, Bay View high was placed in Cluster 3 by Method A, though its racial demographics make it a bit of an edge case there, not too dissimilar from Cluster 1. Then when location is added, it tips the scale so that it is added to its geographically nearby cluster, Cluster 1.

Adding geographic location to the clustering algorithm does not make significant changes to the clusters, though it does move certain edge cases to different clusters, adding or removing them from cluster in their geographic proximity depending on similarity with that cluster. **Therefore, we will select Method A because 1) it is the simpler method, and 2) we are interested in the students served by schools, even if a school serves a student body that is dissimilar from surrounding schools.**

\newpage

```{r}
diff <- ach_clustered %>%
  select(dpi_true_id,
         school_name,
         per_b_aa,
         per_hisp_lat,
         per_white,
         per_asian,
         per_lep,
         per_ed,
         per_swd,
         cluster_nogeo = cluster,
         overall_rating) %>%
  left_join(., ach_clustered_geo %>%
              select(dpi_true_id,
                     cluster_geo = cluster)) %>%
  filter(cluster_nogeo != cluster_geo)

diff %>%
  arrange(school_name) %>%
  select("School" = school_name,
         "Method A" = cluster_nogeo,
         "Method B" = cluster_geo,
         "Black" = per_b_aa,
         "Hisp/Lat" = per_hisp_lat,
         "White" = per_white,
         "Asian" = per_asian,
         "SwD" = per_swd,
         "LEP" = per_lep,
         "ECD" = per_ed) %>%
  modify_at(4:10, percent, .1) %>%
  kbl(booktabs = T, caption = "Schools with Different Clusters",
        align = c("l", rep("c", 9)), linesep = "") %>%
  kable_styling(latex_options = c("striped", "HOLD_position", "scale_down"))

```

```{r fig.height = 7}
diff %>%
  left_join(., geo_schools) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(data = sbds) +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT != 7), inherit.aes = FALSE,
               size = 5, color = "red") +
  geom_sf_text(aes(label = DISTRICT), data = sbds %>% filter(DISTRICT == 7), inherit.aes = FALSE,
               size = 5, nudge_y = -.01,  color = "red") +
  geom_sf() +
  geom_sf_label(aes(label = glue("{school_name}\n{cluster_nogeo} \u2192 {cluster_geo}")), vjust = 1.2,
                color = "white", fill = cfc_darkblue, lineheight = .9) +
  coord_sf(clip = "off") +
  theme_void() +
  labs(fill = "Cluster")
```

# Results

We have already covered a good deal of information regarding the output of Method A. For instance, we have seen that Cluster 3 is the most disadvantaged, Cluster 1 is a bit less disadvantaged, and Cluster 2 is by far the least disadvantaged. This interpretation is supported by the average School Report Card scores for each cluster.

Geographically, the clusters follow the segregation of the city. Cluster 3, which serves a majority of Black students, is clustered on the north side of the city. Cluster 1, which serves a majority of Hispanic/Latino students, is clustered on the south side of the city. Cluster 2 is interesting in that it is geographically dispersed (though with the highest concentration in SBD8 on the southeast side of the city).

Finally, the next several pages provide a list of schools (in alphabetical order) with their assigned clusters, along with their Report Card Scores. All data is from the 2020-21 school year.

\newpage
## List of Schools with Cluster

```{r}
ach_clustered %>%
  arrange(school_name) %>%
  select("School" = school_name,
         "Cluster" = cluster,
         "Asian" = per_asian,
         "Black" = per_b_aa,
         "Hisp/\nLat" = per_hisp_lat,
         "White" = per_white,
         "SwD" = per_swd,
         "LEP" = per_lep,
         "ECD" = per_ed,
         "Overall Score" = overall_score,
         "Ach." = sch_ach,
         "Growth" = sch_growth) %>%
  modify_at(3:9, percent, .1) %>%
  kbl(booktabs = T, caption = "Milwaukee Schools with Clusters",
        align = c("l", rep("c", 11)), linesep = "", longtable = TRUE, ) %>%
  kable_styling(latex_options = c("striped", "hold_position", "repeat_header"),
                font_size = 8) %>%
  column_spec(1, width = "10em") %>%
  column_spec(c(2:12), width = "2em")
```